---
title: "Assignment 3A: Global Baseline Estimate"
Author: Yeimi Perez
format: html

execute:
  echo: true
  output: true
  warning: false
  message: false
---

# Approach

For this assignment, I will use the movie rating survey data collected in the previous assignment to create a recommendation using the Global Baseline Estimate method. Rather than calculating recommendations for every user and every unrated movie, I will focus on one selected rater and one movie that the person has not rated.

My approach will be to first calculate the overall average rating across the movie dataset. I will then calculate the selected user's average rating and the selected movie's average rating. From these values, I will determine the user bias and movie bias and apply the Global Baseline Estimate formula provided in the assignment spreadsheet. One challenge I anticipate is making sure that missing ratings are not included when calculating the averages. I will also need to make sure that I correctly match the selected user's ratings and the selected movie's ratings before calculating the biases.

# Introduction

The purpose of this assignment is to use movie rating data to create a recommendation using the Global Baseline Estimate method. This method uses the overall average rating, along with user and movie rating patterns, to estimate how a user may rate a movie they have not yet rated. For this assignment, I will focus on one selected user and one unrated movie.

```{r}
ratings <- data.frame(
  user_id = c(
    1,1,1,1,1,1,
    2,2,2,2,
    3,3,
    4,4,4,
    5,5,5,
    6,6,6,6
  ),
  
  movie_id = c(
    1,2,3,4,5,6,
    1,2,3,5,
    1,4,
    2,3,4,
    1,3,4,
    3,4,5,6
  ),
  
  rating = c(
    5,4,5,4,4,5,
    4,4,4,5,
    5,5,
    5,4,4,
    4,5,4,
    4,5,4,5
  )
)

ratings
```

ratingsuser_id = which person gave the rating

movie_id = which movie they rated

rating = their score

- 

  ### Calculating the Global Average 

```{r}
global_mean <- mean(ratings$rating)

global_mean
```

### Calculating average for Eduard

```{r}
eduard_ratings <- ratings[
  ratings$user_id == 2,
]

eduard_ratings
```

```{r}
eduard_average <- mean(eduard_ratings$rating)

eduard_average
```

### Calculating Eduard's user bias

```{r}
eduard_bias <- eduard_average - global_mean

eduard_bias
```

This means that Eduard tends to rate movies slightly lower than the overall dataset average.

### Calculating Movie 6' average: **Lion King II: Simba’s Pride**.

```{r}
movie6_ratings <- ratings[
  ratings$movie_id == 6,
]

movie6_ratings
```

Here we isolate everyone who rated movie 6,

```{r}
movie6_average <- mean(movie6_ratings$rating)

movie6_average
```

and we calculate the mean from the two users who rated.

### Calculating Movie's 6 bias

```{r}
movie6_bias <- movie6_average - global_mean

movie6_bias
```

This means that Movie 6 tends to receive higher ratings than the overall dataset average.

### Calculating the Global Baseline Estimate

```{r}
prediction <- global_mean +
  eduard_bias +
  movie6_bias

prediction
round(prediction, 2)
```

This means that Eduard would rate Lion King II: Simba’s Pride approximately 4.8 out of 5.

# Conclusion

For this assignment, I used the Global Baseline Estimate method to predict a rating for one user and one movie that had not been rated. I first calculated the overall average rating, then the selected user's average and bias, followed by the selected movie's average and bias. Using these values, I calculated the predicted rating with the Global Baseline Estimate formula.

The final predicted rating was approximately **4.80 out of 5**. Based on this estimate, the selected movie would be a strong recommendation for this user.
