---
title: "Assignment 3B: Window Functions"
author: "Yeimi Perez"
format:
  html:
    code-fold: false
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# Approach

For Assignment 3B, I will use a synthetic medication adherence time-series dataset containing daily adherence rates across multiple months for diabetes and hypertension medications. The dataset will include the date, month, medication category, and adherence rate, allowing adherence performance to be followed over time and compared month over month.

Using R and `dplyr`, I will group the data by medication category and calculate the required year-to-date average and six-day moving average. I will also use the monthly structure of the dataset to observe changes in adherence performance from one month to the next.

# Introduction

For this assignment, I use a synthetic medication adherence time-series dataset containing daily adherence rates for diabetes and hypertension medications across multiple months. Using R and `dplyr`, I calculate a year-to-date average and six-day moving average for each medication category. The monthly structure of the dataset also allows changes in adherence performance to be observed over time.

```{r}
library(dplyr)
```

### Creating the Synthetic Dataset

```{r}
dates <- seq(
  as.Date("2026-01-01"),
  as.Date("2026-03-31"),
  by = "day"
)
```

```{r}
dates
```

Here we are creating a sequence generated by R starting with January 1, 2026, continued through March 31, 2026, and move forward one day at a time with 90 observations.

```{r}
adherence_data <- data.frame(
  date = rep(dates, 2),
  
  medication_category = rep(
    c("Diabetes", "Hypertension"),
    each = length(dates)
  ),
  
  adherence_rate = c(
    seq(82, 90, length.out = length(dates)),
    seq(86, 93, length.out = length(dates))
  )
)
```

Here we created a full adherence dataset with variables:

|        |                        |
|--------|------------------------|
| `date` | Daily observation date |

|                       |                          |
|-----------------------|--------------------------|
| `medication_category` | Diabetes or Hypertension |

|                  |                                      |
|------------------|--------------------------------------|
| `adherence_rate` | Synthetic daily adherence percentage |

```{r}
adherence_data
```

and created 90 Diab observations and 90 Hypertension observations. Something to note is that the adherence values gradually increase over the three months.

### Adding a Month Variable

```{r}
adherence_data <- adherence_data %>%
  mutate(
    month = format(date, "%B")
  )
```

format(date, "%B") converts the date into the full month name, Jan/Feb/March

```{r}
adherence_data
```

### Arranging the dataset

```{r}
adherence_data <- adherence_data %>%
  arrange(medication_category, date)
```

Here we tell R to order the observations organized by date.

```{r}
adherence_data <- adherence_data %>%
  group_by(medication_category)
```

Here we tell R to group by medication category to perform the upcoming calculations separately for each medication category.This part is essential so that we dont accidentally treat both medication categories as one continuous series.

### Calculating the YTD average

```{r}
adherence_data <- adherence_data %>%
  mutate(
    ytd_average = cummean(adherence_rate)
  )
```

cumean() is the cumulative mean- The average adherence rate from the beginning of the year up through the current date.

```{r}
adherence_data
```

### Calculating the six-day moving average

```{r}
adherence_data <- adherence_data %>%
  mutate(
    six_day_average = rowMeans(
      cbind(
        adherence_rate,
        lag(adherence_rate, 1),
        lag(adherence_rate, 2),
        lag(adherence_rate, 3),
        lag(adherence_rate, 4),
        lag(adherence_rate, 5)
      ),
      na.rm = FALSE
    )
  )
```

lag() is used to look backward at an earlier observation- We take yesterday's adherence rates (adherence_rate, 1) and ask for the adherence rate from five observations earlier (adherence_rate, 5).

```{r}
adherence_data
```

The first five observations read NA under each medication category, this is because we cannot calculate a six-day average until six observations exist.

### Cleaning up the results

```{r}
adherence_data <- adherence_data %>%
  mutate(
    adherence_rate = round(adherence_rate, 2),
    ytd_average = round(ytd_average, 2),
    six_day_average = round(six_day_average, 2)
  )
```

The averages have too many decimals.

```{r}
monthly_summary <- adherence_data %>%
  group_by(medication_category, month) %>%
  summarise(
    monthly_average = mean(adherence_rate),
    .groups = "drop"
  )
```

Here we create a monthly summary to calculate the month-over-month.

```{r}
monthly_summary
```

```{r}
adherence_data <- adherence_data %>%
  mutate(
    month = factor(
      month,
      levels = c("January", "February", "March")
    )
  )
```

Here we organize the months in chronological order

```{r}
monthly_summary <- adherence_data %>%
  group_by(medication_category, month) %>%
  summarise(
    monthly_average = mean(adherence_rate),
    .groups = "drop"
  )
```

### Calculating the actual month-over-month change

```{r}
monthly_summary <- monthly_summary %>%
  group_by(medication_category) %>%
  arrange(month) %>%
  mutate(
    month_over_month_change =
      monthly_average - lag(monthly_average)
  ) %>%
  ungroup()
```

```{r}
monthly_summary
```

For January, the change will be NA because there is no previous month.

### Ungrouping the main dataset

```{r}
adherence_data <- adherence_data %>%
  ungroup()
```

The grouped calculations are all set

```{r}
adherence_data %>%
  select(
    date,
    month,
    medication_category,
    adherence_rate,
    ytd_average,
    six_day_average
  )
```

Displaying the window-function output

```{r}
monthly_summary
```

Displaying the monthly ouput.

The year-to-date average provides a cumulative view of medication adherence performance from the beginning of the year through each date, while the six-day moving average provides a shorter-term view of recent adherence trends. The monthly summary allows adherence performance to be compared across January, February, and March for both medication categories.

# Conclusion

In this assignment, I used a synthetic medication adherence time-series dataset to calculate window functions for diabetes and hypertension medications. The year-to-date average showed cumulative adherence performance, while the six-day moving average provided a view of shorter-term changes in adherence rates.

I also summarized the data by month to compare adherence performance across January, February, and March. This demonstrated how time-series and window-function techniques can be applied to healthcare data to evaluate both long-term and recent performance trends.
