---
title: "Assignment_2A: SQL and R - Movie Ratings"
Author: Yeimi Perez
format: html
editor: visual
---

# Introduction

For this assignment, I created a small movie-rating dataset using ratings collected from six family participants across six selected movies. The purpose of this analysis is to organize the data in a relational PostgreSQL database, use SQL queries to connect and retrieve information from the database, and import the results into R for additional analysis.

```{r}
library(DBI)
library(RPostgres)
```

Loading the packages to communicate with databases. Loading PostgreSQL driver so R connects to PostgreSQL database.

```{r}
con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "Movie Ratings",
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = rstudioapi::askForPassword("Enter PostgreSQL password")
)
```

Here we created an abject called **con** that will store my connection to the database, connect R to my PostgreSQL, and use the driver installed. Then connect to the database names Movie Ratings using designated port and prompt password entry as a pop-up.

```{r}
dbListTables(con)
```

testing the connection

```{r}
ratings_df <- dbGetQuery(
  con,
  "
  SELECT
      u.name,
      m.title,
      r.rating
  FROM ratings r
  JOIN users u
      ON r.user_id = u.user_id
  JOIN movies m
      ON r.movie_id = m.movie_id
  ORDER BY u.name, m.title;
  "
)
```

This creates a new data frame named ratings_df. I asked PostgreSQL to run a SQL query and send the results into R.

```{r}
dbIsValid(con)
```

Verifying the connection is still there.

```{r}
dbGetQuery(con, "SELECT * FROM users;")
```

Verified R can read my table.

# Moving on to R analysis

```{r}
glimpse(ratings_df)
```

Here we are looking at the structure of ratings_df/variables.

```{r}
ratings_df |>
  group_by(title) |>
  summarize(
    average_rating = mean(rating)
  )
```

Here we are grouping all ratings by movie title, then calculate the average rating for each movie.

```{r}
ratings_df |>
  group_by(title) |>
  summarize(
    number_of_ratings = n()
  )
```

Here we are grouping the data by movie, then counting how many ratings each movie received.

```{r}
movie_summary <- ratings_df |>
  group_by(title) |>
  summarize(
    average_rating = mean(rating),
    number_of_ratings = n()
  ) |>
  arrange(desc(average_rating))

movie_summary
```

Here we ask to group ratings by movie, calculate the average and number of ratings, then sort the movies from highest average rating to lowest.

```{r}
user_summary <- ratings_df |>
  group_by(name) |>
  summarize(
    ratings_given = n(),
    average_rating_given = mean(rating)
  )

user_summary
```

Here we group the data by person, count how many movies each person rated, and calculate their average rating.

# Missing Ratings

Not every participant had seen every movie. Rather than assigning a value of zero, which would incorrectly fall outside the 1–5 rating scale, unrated movies were left absent from the `ratings` table. This preserves the distinction between a low rating and a movie that was not rated.

```{r}
movie_summary <- ratings_df |>
  group_by(title) |>
  summarize(
    average_rating = mean(rating),
    number_of_ratings = n()
  ) |>
  arrange(desc(average_rating))

movie_summary
```

```{r}
total_possible <- 6 * 6

actual_ratings <- nrow(ratings_df)

missing_ratings <- total_possible - actual_ratings

missing_ratings
```

Here I'm looking to perform an exploratory data analysis due to the missing ratings, observing a total of 36 possible scores and then counting how many actually exist, with result of 32.

# Conclusion

This assignment introduced the process of collecting data, designing a normalized relational database, querying data with SQL, importing SQL results into R, and producing summary statistics using tidyverse. The resulting dataset resembles the structure of user-item data used in recommender systems and provided practical experience integrating SQL with R.
